---
output: 
  github_document:
    html_preview: false
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```
[![Travis-CI Build Status](https://travis-ci.org/gschofl/biofiles.svg?branch=master)](https://travis-ci.org/gschofl/biofiles)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/gschofl/biofiles?branch=master&svg=true)](https://ci.appveyor.com/project/gschofl/biofiles)
[![CRAN RStudio mirror downloads](http://cranlogs.r-pkg.org/badges/biofiles)](http://cran.r-project.org/web/packages/biofiles/index.html)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/biofiles)](https://cran.r-project.org/package=biofiles)

# biofiles - an interface to GenBank/GenPept files in R

biofiles provides interfacing to GenBank/GenPept or Embl flat
file records. It includes utilities for reading and writing GenBank
files, and methods for interacting with annotation and sequence data.

## Installation

Install the latest stable release of the `biofiles` package from CRAN:

```{r cran-installation, eval = FALSE}
install.packages("biofiles")
```

Install the development version from `github` using the `devtools` package.

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("gschofl/biofiles")
```

## Basic functionality

Let's download a small bacterial genome, _Chlamydophila psittaci_ DC15
(GenBank acc. CP002806; GI 334693792) from NCBI.

```{r chunk1}
# install.packages("reutils")
gb_file <- reutils::efetch("CP002806", "nuccore", rettype = "gbwithparts", retmode = "text")
gb_file
```

Next, we parse the `efetch` object into a `gbRecord` instance.


```{r chunk2}
rec <- biofiles::gbRecord(gb_file)
rec
```


The `summary` function provides an overview over the object:


```{r chunk3}
biofiles::summary(rec)
```

Various getter methods provide access to the data contained in a GenBank record;
for instance:

```{r chunk4}
biofiles::getAccession(rec)
```

```{r chunk5}
biofiles::getGeneID(rec)
```

```{r chunk6}
biofiles::getDefinition(rec)
```

```{r chunk7}
biofiles::getOrganism(rec)
```


```{r chunk8}
biofiles::getSequence(rec)
```

```{r chunk9}
biofiles::getReference(rec)
```


The function `uniqueQualifs()` provides an overview over the feature qualifiers used
in a record:


```{r chunk10}
biofiles::uniqueQualifs(rec)
```

### Genbank Features

The important part of a GenBank record will generally be the list of annotions
or features.

We can access the `gbFeatureList` of a `gbRecord` using `getFeatures()` or `ft()`:


```{r chunk11}
f <- biofiles::ft(rec)
f
```

We can extract features either by numeric subsetting:

```{r chunk12}
f[[1]]
```

or we can subset by feature key:

```{r chunk13}
f["CDS"]
```

A more versatile method to narrow down the list of features of interest is the 
function `filter()`.
For instance, we can filter for all coding sequences (CDS) with the
annotation "hypothetical" in the product qualifiers:

```{r chunk14}
hypo <- biofiles::filter(rec, key = "CDS", product = "hypothetical")
biofiles::summary(hypo)
```

or we can filter for all elongation factors,

```{r chunk15}
elong <- biofiles::filter(rec, key = "CDS", product = "elongation factor")
biofiles::summary(elong)
```

now let's extract the sequence for all elongation factors, and using the tools
from the `Biostrings` packages, translate them into protein sequences. Note, that
in order to do so, we first get the `gbFeatureTable` from the `gbRecord`, as otherwise
we'd just extract the complete sequence associated with the GenBank record.

```{r chunk16}
dna <- biofiles::getSequence(biofiles::ft(elong))
dna
```

```{r chunk17}
str <- biofiles::strand(elong)
dna_revcomp <- c(Biostrings::reverseComplement(dna[str == -1]), dna[str == 1])
aa <- Biostrings::translate(dna_revcomp)
names(aa) <- names(dna_revcomp)
aa
```

We can use the `ranges()` method to extract `GRanges` objects defined in the
Bioconductor package `GenomicRanges`:

```{r chunk18}
elong_ranges <- biofiles::ranges(elong, include = c("locus_tag", "protein_id", "product"))
elong_ranges
```





